A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-10 tidyindex.qmd
more writing in the intro

The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making. However, when analysing specific index examples, we find the composition of indexes is much more complicated than originally expected. In addition to combining multiple variables, indexes can have a spatial and/or temporal component to capture the neighbouring effect in space and the prolonged effect in time. [Constructing indexes from combining variables with spatial and temporal component is further complicated by different treatments to process data from different sources such as satellite imageries, in-situ stations, and large physical models.]

**What is the challenges with current index construction**

While numerous indexes are proposed by researchers, most well-known indexes only use one or two variables. More complex indexes tend to have difficulties to reach consensus given the uncertainty of their behaviors in different scenarios. Typical ways to evaluate new indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e.  crop yield for monitoring agricultural drought.

  - However, more doubt can still raises: Whether the variable used to construct the index is a good set of variables. 
  - Does the index series capture the variables measures. Whether other ways of combining the variables results in difference in the index. 
  - Being able to answer these questions help to prove the robustness, extendibility and dimensionality of the index, as proposed by xxx as criteria for (selecting?) indexes.

**what can be done if people adopt this pipeline/ why it is beneficial?**

  - In order to evaluate indexes, indexes need to be broken down into pieces that it constitutes and then we can analyse how each component will affect the index. 
  - The OECD handbook [@oecd_handbook_2008] provides technical guidelines to construct composite indexes for comparing country performance for policy-making. 
  This paper identifies a more general pipeline framework where statistical indexes (of all kinds) can be built from. 
  Based on the pipeline, an R package is built to validate the indexes, [more details on the functionality here] from understanding multivariate relationships to uncertainty analysis.

<!-- The pipeline approach is general while adaptable to most index construction.  -->
<!-- It allows indexes to be created, studied, and compared in a structured tidy form and enables statistical analysis of indexes to be performed easily:  -->

**who would benefit from this paper**

  - This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes. 
  - Helping index analysts to identify weakness in existing indexes for methodology improvement.

>>>>

The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making. However, when analysing specific index examples, we find indexes are often not just a simple weighted linear aggregation of variables. When indexes have a spatial and/or temporal component, different choices on aggregating space and time will introduce uncertainty. [There is always new variables found to be related to make the decision and should be introduced into the index. The decision made in pre-processing data coming from different sources such as satellite imageries, in-situ stations, and large physical models.] While numerous indexes are proposed in the literature, the well-known ones tend to prefer a simple procedure: they may only consider one or two variables, i.e. SPI or use a decision-tree based process to incorporate multiple variables, i.e. (How El Nino is monitored by major climate institutes). More complex indexes often have difficulties to reach consensus, due to their uncertainty when applied in different scenarios.

**What is the challenges with current index construction**/ **what can be done if people adopt this pipeline/ why it is beneficial?**/**who would benefit from this paper**

Typical ways to evaluate indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e. crop yield for monitoring agricultural drought. While these evaluations show the accuracy of the index to capture, they doesn't address the robustness of the index,[ given variation of the input parameters, or a change of different study areas or across time.] To evaluate the robustness of indexes, indexes need to be broken down into their fundamental building blocks, and then analyse the effect of each component. The OECD handbook [@oecd_handbook_2008] has recommended a set of guidelines for construct composite indexes for measuring country's socio-economic performance. This paper identifies a more general pipeline framework from which indexes can be built from. Based on the pipeline, an R package, tidyindex, is developed to [validate the indexes]. This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes. It also helps index analysts to identify weakness in existing indexes for methodology improvement.

### 2023-07-10 tidyindex.tex
more writing in the intro

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, enhanced, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

The initial plan aims to investigate how variable combinations can
affect the indexes and whether we can use statistical methods to
recommend better combination for decision making. However, when
analysing specific index examples, we find the composition of indexes is
much more complicated than originally expected. In addition to combining
multiple variables, indexes can have a spatial and/or temporal component
to capture the neighbouring effect in space and the prolonged effect in
time. {[}Constructing indexes from combining variables with spatial and
temporal component is further complicated by different treatments to
process data from different sources such as satellite imageries, in-situ
stations, and large physical models.{]}

\textbf{What is the challenges with current index construction}

While numerous indexes are proposed by researchers, most well-known
indexes only use one or two variables. More complex indexes tend to have
difficulties to reach consensus given the uncertainty of their behaviors
in different scenarios. Typical ways to evaluate new indexes is to
compare it with the standard/ well-known indexes, such as SPI or SPEI
for drought indexes, or a response variable relevant for the monitoring
purpose, i.e.~crop yield for monitoring agricultural drought.

\begin{itemize}
\tightlist
\item
  However, more doubt can still raises: Whether the variable used to
  construct the index is a good set of variables.
\item
  Does the index series capture the variables measures. Whether other
  ways of combining the variables results in difference in the index.
\item
  Being able to answer these questions help to prove the robustness,
  extendibility and dimensionality of the index, as proposed by xxx as
  criteria for (selecting?) indexes.
\end{itemize}

\textbf{what can be done if people adopt this pipeline/ why it is
beneficial?}

\begin{itemize}
\tightlist
\item
  In order to evaluate indexes, indexes need to be broken down into
  pieces that it constitutes and then we can analyse how each component
  will affect the index.
\item
  The OECD handbook (OECD, European Union, and Joint Research Centre -
  European Commission 2008) provides technical guidelines to construct
  composite indexes for comparing country performance for policy-making.
  This paper identifies a more general pipeline framework where
  statistical indexes (of all kinds) can be built from. Based on the
  pipeline, an R package is built to validate the indexes, {[}more
  details on the functionality here{]} from understanding multivariate
  relationships to uncertainty analysis.
\end{itemize}

\textbf{who would benefit from this paper}

\begin{itemize}
\tightlist
\item
  This work provides researchers actively developing new indexes with
  tools to evaluate the proposed indexes.
\item
  Helping index analysts to identify weakness in existing indexes for
  methodology improvement.
\end{itemize}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

The initial plan aims to investigate how variable combinations can
affect the indexes and whether we can use statistical methods to
recommend better combination for decision making. However, when
analysing specific index examples, we find indexes are often not just a
simple weighted linear aggregation of variables. When indexes have a
spatial and/or temporal component, different choices on aggregating
space and time will introduce uncertainty. {[}There is always new
variables found to be related to make the decision and should be
introduced into the index. The decision made in pre-processing data
coming from different sources such as satellite imageries, in-situ
stations, and large physical models.{]} While numerous indexes are
proposed in the literature, the well-known ones tend to prefer a simple
procedure: they may only consider one or two variables, i.e.~SPI or use
a decision-tree based process to incorporate multiple variables,
i.e.~(How El Nino is monitored by major climate institutes). More
complex indexes often have difficulties to reach consensus, due to their
uncertainty when applied in different scenarios.

\textbf{What is the challenges with current index construction}/
\textbf{what can be done if people adopt this pipeline/ why it is
beneficial?}/\textbf{who would benefit from this paper}

Typical ways to evaluate indexes is to compare it with the standard/
well-known indexes, such as SPI or SPEI for drought indexes, or a
response variable relevant for the monitoring purpose, i.e.~crop yield
for monitoring agricultural drought. While these evaluations show the
accuracy of the index to capture, they doesn't address the robustness of
the index,{[} given variation of the input parameters, or a change of
different study areas or across time.{]} To evaluate the robustness of
indexes, indexes need to be broken down into their fundamental building
blocks, and then analyse the effect of each component. The OECD handbook
(OECD, European Union, and Joint Research Centre - European Commission
2008) has recommended a set of guidelines for construct composite
indexes for measuring country's socio-economic performance. This paper
identifies a more general pipeline framework from which indexes can be
built from. Based on the pipeline, an R package, tidyindex, is developed
to {[}validate the indexes{]}. This work provides researchers actively
developing new indexes with tools to evaluate the proposed indexes. It
also helps index analysts to identify weakness in existing indexes for
methodology improvement.
